---
title: "Lecture 15 - GPs for GLMs and Spatial Data"
author: "Ethan Shen"
date: "5/26/2020"
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)
library(purrr)

ggplot2::theme_set(ggplot2::theme_bw())

source("/Users/ethanshen/Documents/Projects for Fun /Spatio-Temporal Modeling/util.R")
knitr::opts_chunk$set(
  warning=FALSE,
  message=FALSE
)
```

# Logistic Model Example 

```{r}
set.seed(20180308)
n=200

eta = function(x) 2.5*sin(2.1*pi*(x-0.25)) 

d = data_frame(x=runif(n)) %>%
  mutate(
    eta = eta(x),
    prob = inv_logit(eta(x)),
    y = rbinom(length(x), size=1, prob = inv_logit(eta(x)))
  )

(ggplot(d, aes(x=x, y=eta)) + geom_line() + labs(title="Untransformed") + 
    ggplot(d, aes(x=x, y=prob)) + geom_line() + labs(title="Transformed"))  / 
  (ggplot(d, aes(x=x, y=y)) + geom_jitter(height = 0.05))
```

### JAGS

```{r}
logistic_model = "model{
  for(i in 1:length(y)) {
    y[i] ~ dbern(p[i])
    logit(p[i]) = beta0 + w[i]
  }

  w ~ dmnorm(rep(0,length(y)), inverse(Sigma)) # GP, mean 0 multivariate procses

  for (i in 1:(length(y)-1)) {
    for (j in (i+1):length(y)) {
    
    # exponential covariance
    
      Sigma[i,j] = sigma2 * exp(- l * d[i,j]) 
      Sigma[j,i] = Sigma[i,j]
    }
  }

  for (i in 1:length(y)) {
    Sigma[i,i] = sigma2
  }  
  
  beta0 ~ dnorm(0, 1)
  sigma2 = 1/tau
  tau ~ dgamma(1, 2)
  l ~ dunif(3/0.5, 3/0.01)
}"

if (!file.exists("jags_logistic_model.rds")) {
  m_JAGS = rjags::jags.model(
    textConnection(logistic_model),
    data = list(
      y = (d$y),
      d = dist(d$eta) %>% as.matrix()
    )
  )
  
  update(m_JAGS, n.iter=5000)
  
  JAGS_samps = rjags::coda.samples(
    m_JAGS, variable.names=c("beta0", "l", "sigma2"),
    n.iter=10000
  )
  saveRDS(JAGS_samps, file="jags_logistic_model.rds")
} else {
  JAGS_samps = readRDS("jags_logistic_model.rds")
}
```

This takes too long to sample with JAGS. 

### JAGS Model Diagnostics 

```{r}
tidybayes::gather_draws(JAGS_samps, beta0, l, sigma2) %>%
  filter(row_number() %% 10 == 1) %>%
  mutate(.iteration = floor(.iteration / 10) + 1) %>%
  ggplot(aes(x = .iteration, y = .value, color = .variable)) + 
  geom_line() + 
  facet_grid(.variable~., scales='free_y') +
  guides(color=FALSE)
```

### spBayes

```{r}
y = d$y
x = d$x

coords = cbind(x, 0)
n = nrow(coords)

n.batch = 200
batch.length = 100

n.samples = n.batch * batch.length
burnin = n.samples*0.5 + 1
thin = 10

if (!file.exists("logistic_model.rds")) {
  m_spBayes = spBayes::spGLM(
    y~1, family="binomial", coords=coords,
    starting=list("beta"=0, "phi"=3/0.053,"sigma.sq"=1, "w"=0),
    tuning=list("beta"=0.5, "phi"=0.5, "sigma.sq"=0.5, "w"=0.5),
    priors=list("beta.Normal"=list(0,1), "phi.Unif"=c(3/0.5, 3/0.01), "sigma.sq.IG"=c(2, 1)),
    amcmc=list("n.batch"=n.batch, "batch.length"=batch.length, "accept.rate"=0.43),
    cov.model="exponential", verbose=TRUE, n.report=10)
  
  saveRDS(m_spBayes, file="logistic_model.rds")
} else {
  m_spBayes = readRDS("logistic_model.rds")
}
```

### spBayes Model Diagnostics 

```{r}
mcmc = m_spBayes$p.beta.theta.samples %>% 
  window(start = burnin, thin = thin) %>%
  tidybayes::gather_draws(`(Intercept)`, phi, sigma.sq)

ggplot(mcmc, aes(x=.iteration, y=.value, color=.variable)) +
  geom_line() +
  facet_grid(.variable~., scales = "free_y") +
  guides(color=FALSE)
```

phi is range parameter 

### spBayes Model Fit 

```{r}
pred = spBayes::spPredict(m_spBayes, pred.coords = coords, pred.covars = matrix(rep(1,n)), start = burnin, thin=thin, verbose = FALSE)

pred = cbind(
  t(pred$p.w.predictive.samples),
  t(pred$p.y.predictive.samples)
) 
colnames(pred) = c(paste0("eta[",1:n,"]"), paste0("p[",1:n,"]"))

tidybayes::gather_draws(pred, eta[i],p[i]) %>%
  full_join(
    select(d, x) %>% mutate(i = 1:n()),
    by = "i"
  ) %>%
  ungroup() %>%
  mutate(.variable = case_when(
    .variable=="p" ~ "prob",
    TRUE~as.character(.variable))
  ) %>%
  ggplot(aes(x=x, y=.value)) +
  tidybayes::stat_lineribbon(alpha=0.5, color="red") +
  geom_line(data=tidyr::gather(d, .variable, .value, -x, -y), size=2, linetype=2) +
  facet_wrap(~.variable, scales = "free_y")
```

In the original data, there are a couple of points that are near 0 for `x` = 0.5. This is why the red line dips in the `prob` plot, which drags down the plot. Also, exp covariance function doesn't have to be smooth.

## Model vs GLM

```{r}
g = glm(y~poly(x,2), data=d, family="binomial")
```

```{r echo=FALSE}
rbind(
  mutate(d, type = "truth"),
  mutate(d, 
         type = "glm prediction",
         eta = predict(g, type="link"),
         prob = predict(g, type="response")
  )
) %>% 
  tidyr::gather(term, estimate, -x, -y, -type) %>%
  ggplot(aes(x=x, y=estimate, color=type, linetype=type)) +
  geom_line(size=2) +
  facet_wrap(~term, scales = "free_y") +
  scale_color_manual(values = c("red","black"))
```

This GLM has to be smooth because we are fitting a quadratic GLM. However, it can't flatten out at the tails. 

# Spatial Data 

```{r}
library(sf)
nc  = st_read("data/nc_counties.gpkg", quiet=TRUE, stringsAsFactors=FALSE)
air = st_read("data/airports.gpkg", quiet=TRUE, stringsAsFactors=FALSE)
hwy = st_read("data/us_interstates.gpkg", quiet=TRUE, stringsAsFactors=FALSE)
```

```{r}
# for each county, there is a multipolygon that contains the boundaries
# geometry: multipolygon
head(tbl_df(nc))

# geometry: point
head(tbl_df(air))

# geometry: multilinestring
head(tbl_df(hwy))
```

## sf Classes

```{r}
str(nc)
```

`sf` are also data frames. The geometry column is of type `sfc`, list of items. `sfg` are the actual entries, that contain the geometries. 

## Projections

```{r}
st_crs(hwy) ; st_crs(air)
par(mar=c(3,3,3,0.1), mfrow=c(1,3))
plot(st_geometry(nc),  axes=TRUE, main="nc")
plot(st_geometry(air), axes=TRUE, pch=16, col="blue", main="air")
plot(st_geometry(hwy), axes=TRUE, col="red", main="hwy")
```



## Example 1 - What counties are adjacent to Durham County? 

```{r}
# one result for every entry/county in NC
st_touches(
  nc,
  filter(nc, COUNTY == "Durham County")
) %>% str()

# one result 
neighbors = st_touches(
  filter(nc, COUNTY == "Durham County"),
  nc
) %>%
  unlist() %>%
  slice(nc, .)
```

### Base R Plots

```{r}
plot(st_geometry(nc))
plot(st_geometry(neighbors), add=TRUE, col="lightblue")
plot(st_geometry(nc[nc$COUNTY == "Durham County",]), add=TRUE, col="red")
```

### ggplot 

```{r}
ggplot() + 
  geom_sf(data=nc, fill="white") + 
  geom_sf(data=nc %>% filter(COUNTY=="Durham County"), fill="red") + 
  geom_sf(data=neighbors, fill = "lightblue")
```

## Example 2 - Which counties have more than 4 neighbors? 

```{r}
lengths = st_touches(nc) %>% # for each county, tells you which counties it neighbors 
  map_int(length)   # calculates how many neighbors for each county

many_neighbors = which(lengths > 4) # gives us indices of counties with more than 4 neighbors

ggplot() + 
  geom_sf(data=nc, fill="white") + 
  geom_sf(data=nc[many_neighbors, ], fill = "lightblue")
```

## Example 3 - Which counties have an airport? 

```{r}
# most of these will be 0, but we want to find ones where length > 0
airport_counties = st_intersects(nc, air) %>% map_int(length) %>% {which(. > 0)}

ggplot() + 
  geom_sf(data=nc, alpha = 0) + 
  geom_sf(data=nc[airport_counties, ], fill = "lightblue") + 
  geom_sf(data=air %>% filter(STATE=="NC"), color = "red") 
```

### Base R plot 

```{r}
plot(st_geometry(nc))
plot(st_geometry(nc[airport_counties,]), add=TRUE, col="lightblue")
plot(st_geometry(air), add=TRUE, col="red", pch=16)
```

Issue is that it is giving us airports that are outside NC. 

```{r}
# giving ONLY airports inside NC 
nc_airports = st_intersection(nc, air)

airport_counties = st_intersects(nc, nc_airports) %>% map_int(length) %>% {which(. > 0)}

plot(st_geometry(nc))
plot(st_geometry(nc[airport_counties,]), add=TRUE, col="lightblue")
plot(st_geometry(nc_airports), add=TRUE, col="red", pch=16)
```





