---
title: "Lecture 16 - Spatial Data and Cartography"
author: "Ethan Shen"
date: "5/30/2020"
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F,
                      message=F)
library(tidyverse)
library(sf)
library(sp)
ggplot2::theme_set(new=theme_bw())
```

# NC Data 

```{r}
nc = st_read(system.file('shape/nc.shp', package="sf"))
```

### Plotting - Base R

Discrete variable = discrete color palette 

These are plotting `sf` objects. 

```{r}
plot(nc, max.plot=14)
plot(nc[,"BIR74"], 
     graticule = st_crs(nc), # coordinate reference system of the NC dataset 
     axes=T)
```

This plots an `sfc` object. 

```{r}
plot(st_geometry(nc), axes=TRUE)
```

### Plotting - ggplot2

Automatically convert to latitude/longitude projection

```{r}
ggplot(nc) + 
  geom_sf(aes(fill=SID79)) +
  scale_fill_viridis_c()
```

# Meuse Data (river data)

`meuse` is actual observed data (measurement of heavy metal at a specific point (x, y coordinate))

`meuse.riv` is polygon describing boundary of river 

```{r}
data(meuse, meuse.riv)

meuse = st_as_sf(meuse, # converting sp to sf
                 coords=c("x", "y"), # need to also specify coordinates 
                 crs=28992 # coordinate reference system 
)  

meuse.riv = st_polygon(list(meuse.riv)) %>% # turning initial matrix object of x's and y's into a polygon (polygon has shell. first argument in st_polygon is the external bit, internal bits (additional arguments/additional matrices) would be holes. a river doesn't have holes)
  st_sfc() %>% # turning sfg object into sfc
  st_set_crs(28992) # geometries don't have reference systems, but sfc's do 

head(tbl_df(meuse))
head(tbl_df(meuse.riv))
```

```{r}
ggplot() +
  geom_sf(data=meuse, aes(color=lead)) + 
  geom_sf(data=meuse.riv, fill="lightblue", col=NA)

# tells us bounding box of our data 
# st_bbox(meuse)

ggplot() +
  geom_sf(data=meuse.riv, fill="lightblue", col=NA) +
  geom_sf(data=meuse, aes(color=lead)) + 
  ylim(st_bbox(meuse)["ymin"], st_bbox(meuse)["ymax"])
```

# Geometry Manipulation 

### Casting 

changing type 

```{r}
nc_pts = st_cast(nc, "MULTIPOINT") # changing multipolygon --> multipoint (which is just coordinates of boundaries of each county)

plot(st_geometry(nc_pts), pch=16, cex=0.5)

nc_p = st_cast(nc, "POINT") # gives individual points 

plot(st_geometry(nc_p), pch=16, cex=0.5) # plot is the same thing 
```

### Grouping Features 

```{r}
union = st_union(nc)
tbl_df(union)
plot(union)
```

`st_union` doesn't know how to group other columns, so it just deletes them.

```{r}
nc_cut = nc %>%
  mutate(
    # longitude of centroid of each county 
    X = st_centroid(nc) %>% 
      st_coordinates() %>% # matrix of X and Ys 
      .[,1],
    # latitude of centroid of each county 
    Y = st_centroid(nc) %>% 
      st_coordinates() %>% # matrix of X and Ys 
      .[,2],
    region = cut(X, breaks=5))

ggplot(nc_cut) + 
  geom_sf(aes(fill=region))

nc_cut %>%
  group_by(region) %>%
  summarise() %>%  # will union geometries together, only end up with geometries
  ggplot() + 
  geom_sf(aes(fill=region))
```

# NC Highway Example 

```{r}
hwy_nc = st_intersection(hwy %>% st_transform(st_crs(nc)), # not transforming CRS of hwy will cause error
                         nc)
ggplot() +
  geom_sf(data=nc) +
  geom_sf(data=hwy_nc, col="red")
```

### Changing Projection to UTM 

```{r}
nc_utm = st_transform(nc, "+proj=utm +zone=17 +datum=NAD83 +units=m +no_defs")
ggplot() +
  geom_sf(data=nc_utm) + 
  geom_sf(data=hwy_nc, col='red')
```

### Adding Buffer 

```{r}
hwy_nc_buffer = hwy_nc %>%
  st_transform("+proj=utm +zone=17 +datum=NAD83 +units=m +no_defs") %>% 
  st_buffer(10000) # specified to be in meters

hwy_nc_buffer_km = hwy_nc %>%
  st_transform("+proj=utm +zone=17 +datum=NAD83 +units=km +no_defs") %>% 
  st_buffer(10) # specified to be in kilometers

ggplot() +
  geom_sf(data=nc_utm) +
  geom_sf(data=hwy_nc, color='red') + 
  geom_sf(data=hwy_nc_buffer, fill='red', alpha=0.3)

ggplot() +
  geom_sf(data=nc_utm) +
  geom_sf(data=hwy_nc, color='red') + 
  geom_sf(data=hwy_nc_buffer_km, fill='red', alpha=0.3)
```

Why are there little circles showing up in the plot? The interstates are multiple linestrings, not one "continuous one", which is why there are segments. 

We want to fix this by making just one large polygon. 

### Unioning Buffer 

```{r}
# one polygon
hwy_nc_buffer_union = hwy_nc_buffer %>%
  st_union() %>%
  st_sf()

ggplot() +
  geom_sf(data=nc_utm) +
  geom_sf(data=hwy_nc, color='red') + 
  geom_sf(data=hwy_nc_buffer_union, fill='red', alpha=0.3)
```

# Example - Counties within 10 km of an interstate highway

```{r}
ten_miles_within = st_intersects(nc, hwy_nc_buffer_union %>% 
                                   st_transform(st_crs(nc))) %>% 
  purrr::map_int(length) %>% 
  {which(. > 0)}

ggplot() +
  geom_sf(data=nc, fill=NA) +
  geom_sf(data=nc[ya,], fill="lightblue") + 
  #geom_sf(data=hwy_nc, color='red') + 
  geom_sf(data=hwy_nc_buffer_union, fill='red', alpha=0.3) + 
  labs(caption = paste0(length(ten_miles_within), " counties within 10 km of an interstate"))
```

# Gerrymandering Example 

```{r}
nc_house = st_read("data/nc_districts112.gpkg") %>%
  st_transform("+proj=utm +zone=17 +datum=NAD83 +units=km +no_defs")
plot(nc_house[,"DISTRICT"], axes=TRUE)
```

We want to measure some of this craziness in boundaries. These are obviously not geographically compact. 

### Reock Score 

Measure of compactness calculated as the ratio area of a shape to the area of its minimum bounding circle. Closer to 1 is good 

```{r}
circs = nc_house %>%
  st_geometry() %>%
  lwgeom::st_minimum_bounding_circle()

ggplot() + 
  geom_sf(data=nc_house, aes(fill=DISTRICT)) + 
  geom_sf(data=circs, fill=NA)
```

### Calculating Reock Score 

```{r}
nc_house = nc_house %>%
  mutate(reock = st_area(nc_house) / st_area(circs)) # calculating area of districts + circles

plot(nc_house[, "reock"])
```

# Raster Data 

```{r}
meuse_rast = raster::raster(system.file('external/test.grd', package="raster"))

# colors are actual values, white is missing data / NAs
plot(meuse_rast)
plot(meuse.riv, add=TRUE, col="lightblue")
```

# Spatial Models 

```{r}
head(st_coordinates(meuse))
```






 