---
title: "Lecture 4 - GLMs + Residual Analysis"
author: "Ethan Shen"
date: "5/17/2020"
output: 
  html_document:
    keep_md: TRUE
---

```{r setup message=FALSE, warning=FALSE}
library(tidyverse)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r warning=FALSE}
load("data/anguilla.Rdata")
set.seed(20180130)

part = modelr::resample_partition(anguilla, c(train=0.75, test=0.25))

anguilla = as.data.frame(part$train) %>% as_tibble()
anguilla_test = as.data.frame(part$test) %>% as_tibble()
```

# Simple Model 

```{r}
g = glm(presence ~ SegSumT, family=binomial, data=anguilla)
summary(g)
```

### Model Fit 

```{r}
inv_logit = function(x) 1/(1+exp(-x))

d_g = anguilla %>%
  mutate(p_hat = predict(g, anguilla, type="response"))

d_g_pred = data.frame(SegSumT = seq(11,25,by=0.1)) %>% 
  modelr::add_predictions(g,"p_hat") %>%
  mutate(p_hat = inv_logit(p_hat))
```

```{r}
ggplot(anguilla, aes(x=SegSumT, y=presence)) +
  geom_point(alpha=0.5) +
  geom_line(data=d_g_pred, aes(y=p_hat), color="red")
```

Not doing great. 

### Separation

```{r}
ggplot(d_g, aes(x=p_hat, y=presence, color=as.factor(presence))) +
  geom_point(alpha=0.5) +
  labs(color="presence")
```

### Residuals 

```{r}
d_g = d_g %>%
  mutate(resid = resid(g))

ggplot(d_g, aes(x = SegSumT, y=resid, color=as.factor(presence))) + 
  geom_point(alpha=0.3) + 
  geom_hline(yintercept = 0, color="red")
```

A normal residual plot isn't super informative for GLMs. 

### Binned Residuals 

```{r}
d_g %>%
  mutate(bin = p_hat - (p_hat %% 0.01)) %>%
  group_by(bin) %>%
  summarize(resid_mean = mean(resid)) %>%
  ggplot(aes(y=resid_mean, x=bin)) +
  geom_point() + 
  geom_hline(yintercept = 0, color="red")
```

### Pearson Residuals 

```{r}
d_g = d_g %>% mutate(pearson = (presence - p_hat) / (p_hat * (1-p_hat)))
ggplot(d_g, aes(x=p_hat, y=pearson, color=as.factor(presence))) + 
  geom_point(alpha=0.3) + 
  geom_hline(yintercept = 0, color="red")
```

### Binned Pearson Residuals 

```{r}
d_g %>%
  mutate(bin = p_hat - (p_hat %% 0.01)) %>%
  group_by(bin) %>%
  summarize(pearson_mean = mean(pearson)) %>%
  ggplot(aes(y=pearson_mean, x=bin)) +
  geom_point()
```

### Deviance Residuals 

```{r}
d_g = d_g %>%
  mutate(deviance = sign(presence - p_hat) *sqrt(-2 * (presence*log(p_hat) + (1-presence)*log(1 - p_hat) )))
ggplot(d_g, aes(x=p_hat, y=deviance)) +
  geom_point(alpha=0.1) 
sum(d_g$deviance^2)
```

### Binned Deviance Residuals

```{r}
d_g %>%
  mutate(bin = p_hat - (p_hat %% 0.01)) %>%
  group_by(bin) %>%
  summarize(deviance_mean = mean(deviance)) %>%
  ggplot(aes(y=deviance_mean, x=bin)) +
  geom_point()
```

## Binned Residuals All Together

```{r}
d_g %>%
  select(p_hat:deviance) %>%
  mutate(bin = p_hat - (p_hat %% 0.01)) %>%
  tidyr::gather(type, value, -p_hat, -bin) %>%
  mutate(type = as_factor(type)) %>%
  group_by(type, bin) %>%
  summarize(mean = mean(value)) %>%
  ggplot(aes(x = bin, y = mean, color = type)) + 
  geom_point() + 
  facet_wrap(~type, ncol=1, scale="free_y")
```

Pearson and Deviance residuals are fine, should not use regular residuals. 

# Full Model 

```{r}
f = glm(presence~., family=binomial, data=anguilla)
summary(f)
```

### Residuals vs. Fitted 

```{r}
d_f = anguilla %>%
  mutate(p_hat = predict(f, anguilla, type="response")) %>%
  mutate(
    resid = presence - p_hat,
    pearson = (presence - p_hat) / (p_hat * (1-p_hat)),
    deviance = sign(presence - p_hat) *sqrt(-2 * (presence*log(p_hat) + (1-presence)*log(1 - p_hat) ))
  ) %>%
  mutate(bin = p_hat - (p_hat %% 0.025))

d_f %>%
  select(p_hat:bin) %>%
  tidyr::gather(type, value, -p_hat, -bin) %>%
  mutate(type = as_factor(type)) %>%
  group_by(type, bin) %>%
  summarize(mean = mean(value)) %>%
  ggplot(aes(x=bin, y=mean, color=type)) +
  geom_point() +
  facet_wrap(~type, ncol=1, scale="free_y")
```

## Model Performance

```{r}
ggplot(d_f, aes(x=p_hat, y=presence, color=as.factor(presence))) +
  geom_point(alpha=0.5) +
  labs(color="presence", title="Full Model")
```

### ROC/AUC

```{r}
res_obs = bind_rows(
  select(d_g, presence, p_hat) %>% mutate(model = "SegSumT Model"),
  select(d_f, presence, p_hat) %>% mutate(model = "Full Model")
)

gg_roc = ggplot(res_obs, aes(d=presence, m=p_hat, color=model)) +
    plotROC::geom_roc(n.cuts=0) +
    plotROC::style_roc()

auc = plotROC::calc_auc(gg_roc)["AUC"]
   
gg_roc +  
  annotate(
    "text", x = .75, y = .25, 
    label = paste("AUC =", round(auc, 3))
  )
```

### Out of Sample Predictive Performance

```{r}
full_res = anguilla_test %>% 
  modelr::add_predictions(g,"SegSumT Model") %>%
  modelr::add_predictions(f,"Full Model") %>%
  select(presence, "SegSumT Model","Full Model") %>%
  tidyr::gather(model, p_hat, -presence) %>%
  mutate(p_hat = inv_logit(p_hat)) %>%
  mutate(type = "Test") %>%
  bind_rows(
    res_obs %>% mutate(type = "Train"),
    .
  ) %>%
  mutate(
    model = forcats::as_factor(model),
    type = forcats::as_factor(type)
  )

gg_roc = ggplot(full_res, aes(d=presence, m=p_hat, color=model, linetype=type)) +
    plotROC::geom_roc(n.cuts=0) +
    plotROC::style_roc()

auc = plotROC::calc_auc(gg_roc)["AUC"]
   
gg_roc +  
  annotate(
    "text", x = .75, y = .25, 
    label = paste("AUC =", round(auc, 3))
  )
```






